Semantic Preserving Bijective Mappings for Expressions involving Special Functions between Computer Algebra Systems and Document Preparation Systems

### 2018-12-03 main.tex
minor wording change
Fix 'hence' in text (see #22)

Hence, a \gls*{cas} user needs to fully understand the properties and special definitions (such as the position of branch cuts) in the \gls*{cas} to avoid mistakes during a translation~\parencite{Maple:Cuts}. Hence, a manual translation process is not only laborious, but also prone to errors. Note that this general problem has been named as automatic \gls*{p2c} conversion~\parencite{POM-Tagger}.

@@

In addition, we also introduce the notion of \textit{direct} translations. Most mathematical objects in one system have a direct counterpart in other systems. However, later in the paper, we will explain that a translation from one specific mathematical object to its counterpart in the other system is not always \textit{appropriate}. Also, not every mathematical object has a counterpart in other systems. We call a translation to its counterpart \textit{direct}. For example, the translation~(\ref{eq:cos-def}) is \textit{direct}, while a translation to the definition of the cosine function
\begin{equation*}
\verb|\cos@{z}| \overset{\langMaple}{\mapsto} \verb|(exp(I*z)+exp(-I*z))/2|
\end{equation*}
is not a \textit{direct} translation even though it is {\it appropriate}.
Note that partial results of this paper have been published in~\parencite{CICM:Paper}.

\section{Related Work}\label{sec:related-work}
Since \LaTeX{} became the de facto standard for writing papers in mathematics, most \gls*{cas} provide simple functions to import and export mathematical \LaTeX{} expressions\footnote{The selected \gls*{cas} Maple, Mathematica, Matlab, and SageMath provide import and/or export functions for \LaTeX:\\
Maple, \url{http://www.maplesoft.com/support/help/Maple/view.aspx?path=latex} seen 06/2017;
Mathematica, \url{https://reference.wolfram.com/language/tutorial/GeneratingAndImportingTeX.html} seen 06/2017;
Matlab, \url{https://www.mathworks.com/help/symbolic/latex.html} seen 06/2017;
SageMath, \url{http://doc.sagemath.org/html/en/tutorial/latex.html} seen 06/2017.}. Those tools have two essential problems. They are only able to import simple mathematical expressions, where the semantics are unique. For example, the internal \LaTeX{} macro \verb|\frac| always indicates a fraction. However, for more complex expressions, e.g., the Jacobi polynomial in \Cref{tab:JacobiP-usecase}, the import functions fail. The second problem appears in the export tools. Mathematical expressions in \gls*{cas} are fully semantic. Otherwise the \gls*{cas} wouldn't be able to compute or evaluate the expressions. During the export process, the semantic information is lost, because generic \LaTeX{} is not able to carry sufficient semantic information. Because of these problems, an exported expression cannot be imported to the same system again in most cases (except for simple expressions such as those described above). Our tool attempts to solve these problems and provide round-trip translations between \LaTeX{} and \gls*{cas}.

@@

\vspace{-0.2cm}
The position of the branch cut of the arccotangent function differs after the {\it direct} translation~\eqref{eq:acot-alternatives}, which may lead to incorrect calculations later on. The alternative translations~\eqref{eq:acot-alternatives-1} and~\eqref{eq:acot-alternatives-2} use other functions instead of the arccotangent function. The arctangent function~\eqref{eq:acot-alternatives-1} and the natural logarithm~(\ref{eq:acot-alternatives-2}) have the same positioned branch cuts as in the \gls*{dlmf} and in \Maple. Hence, translation~(\ref{eq:acot-alternatives-1}) solves this issue as long as the user does not evaulate the function at $z = 0$, while translation~(\ref{eq:acot-alternatives-2}) solves the issue except at $z=-\iunit$. Note that none of the translations~(\ref{eq:acot-alternatives}-\ref{eq:acot-alternatives-2}) are {\it appropriate}.

@@

The parsed tree generated by the \gls*{pom}-Tagger is not a mathematical expression tree. The \gls*{pom} project aims to disambiguate mathematical \LaTeX{} expressions and generates an expression tree. However, in the current state, many expressions still cannot be disambiguated. Therefore, the \gls*{pom}-tagger generates a raw parsed tree where each token in the \LaTeX{} expression is a node in the tree. We call this parsed tree the \gls*{pom-pt}.

@@

With these tools, we can translate a \LaTeX{} expression by translating the \gls*{pom-pt} node by node and perform group or reordering operations for some special cases. The algorithm is realized in a simple recursive structure. Whenever the algorithm finds a leaf, it can translate this single term. If the node is not a leaf, it starts to translate all children of the node recursively. This idea appears to be a practical and elegant solution. However, it has a significant drawback. It cannot be used to translate functions. Since the arguments of functions are following siblings in the \gls*{pom-pt}, the algorithm needs to lookahead when a leaf is a known function, e.g., in the case of a semantic macro with arguments (see~\Cref{fig:syntax-tree-usecase}). Algorithm~\ref{alg:translation} is an improved version with lookahead functionality.

@@

Since it is more convenient to parse expressions using uniform notations, most programming 
languages (and \gls*{cas} as well) internally use prefix or postfix notation and do not mix 
the notations in one expression.
However, the common practice in science is to use mixed notations in expressions. Since the \gls*{pom} has rarely implemented mathematical grammatical rules, it takes the input as it is and does not build an expression tree. Therefore, it parses all four examples from~\Cref{tab:notations} to four different \gls*{pom-pt}s rather than to one unique expression tree. In general, this is not a problem for our translation process since most \gls*{cas} are familiar with most common notations. Therefore, the translator does not need to know that `$a$' and `$b$' are the operands of the binary operator `$+$' in `$a+b$.' The translator could simply translate the symbols in `$a+b$' in the same order as they appear in the expression and the \gls*{cas} would understand it. However, there are two new problems with this approach.
\begin{enumerate}\item \label{prob:1} The translated expression is only syntactically correct if the input expression was syntactically correct.
\item \label{prob:2} We cannot translate expressions to \gls*{cas} which use non-standard notations.
\end{enumerate}

@@

\subsection{Subtranslators}\label{sec:subtranslators}
The \verb|SequenceTranslator| translates the \textit{sequence} and \textit{balanced expressions} in the \gls*{pom-pt}. If a node $n$ is a leaf and the represented symbol is an open bracket (parentheses, square brackets and so on), the following nodes are also taken as a \textit{sequence}. Hence, combined with the recursive translation approach, the 
\verb|SequenceTranslator| 
also checks balancing of parentheses in expressions. An expression such as `$(a]$' produces a mismatched parentheses error. On the other hand, this is a problem for real interval expressions such as `$[a,b)$'. In the current version, the program cannot distinguish between mismatched parentheses and half-opened, half-closed intervals. Whether an expression is an interval or another expression is difficult to decide and can depend on the context. Also, the parentheses checker could simply be deactivated to allow mismatched parentheses in an expression. Another option is to use interval macros. e.g., \verb|\intcc@{a}{b} = [a,b]|.

@@

However, there are two essential problems with this approach. Testing whether expressions are {\it appropriate} representations of each other is a challenging task for \gls*{cas} and they often have difficulties testing simple equations symbolically. For example, 
consider
\parencite[(4.35.34)]{NIST:DLMF}
\begin{equation*}
\sinh \left( x+\iunit y \right) = \sinh x \cos y + \iunit \cosh x \sin y,
\end{equation*}
as a difference of the left- and right-hand sides cannot be simplified to zero by default. Furthermore, this approach only checks forward translations because there is no way to automatically check whether two \LaTeX{} expressions are {\it appropriate} or {\it inappropriate} representations of each other (again this could become feasible with our translator). We use \Maple's \texttt{simplify} function to check if the difference of the left-hand side and the right-hand side of the equation is equal to zero. In addition, we use \texttt{simplify} and check if the division of the right-hand side by the left-hand side returns a numerical value or not. The simplification function is the most powerful function to check whether two expressions are {\it appropriate} representations in \Maple. However, there are several cases where simplification fails. Because of implementation details, there are some techniques that help \Maple{} to find possible simplifications. For example, we can force \Maple{} to convert the formula
\begin{equation}
\sinh{x} + \sin{x}
\end{equation}
to an {\it appropriate} representation using their exponential representations, namely
\begin{equation}
\frac{1}{2}\expe^x - \frac{1}{2}\expe^{-x} - \frac{1}{2} \iunit \left( \expe^{\iunit x}-\expe^{-\iunit x} \right).
\end{equation}
With such pre-conversions, we are able to improve the simplification process in \Maple. However, the limitations of the \texttt{simplify} function are still the weakest part of this verification approach. Consider the complex example~\parencite[(12.7.10)]{NIST:DLMF}
\begin{equation}\label{eq:branch-cut-near}
\displaystyle U(0,z) = \sqrt{\frac{z}{2\cpi}} K_{\frac{1}{4}}\left(\frac{1}{4}z^2\right),
\end{equation}
where $U(0,z)$ is the parabolic cylinder function and $K_\nu(z)$ is the modified Bessel function of the second kind. Both functions are well-defined in both systems and we can define a \textit{direct} translation for~\eqref{eq:branch-cut-near}. 
The modified Bessel function of the second kind has its branch cut in \Maple{} and in the \gls*{dlmf} at $z < 0$. However, the argument of $K$ contains a $z^2$. If $|\ph{z}| \in \left(\frac{\cpi}{2}, \cpi\right)$ the value of the right-hand side of~\eqref{eq:branch-cut-near} would be no longer on the principal branch. However, \Maple{} will still compute the principal values independently of the value of $z$. Hence, a translation
\begin{equation}
\verb|\BesselK{\frac{1}{4}}@{\frac{1}{4}z^2}| \overset{\langMaple}{\mapsto} \verb|BesselK(1/4,(1/4)*z^2)|
\end{equation}
is incorrect if $|\ph{z}| \in \left(\frac{\cpi}{2}, \cpi\right)$ and one should instead use the analytic continuation for the right-hand side of (\ref{eq:branch-cut-near}). 
To evaluate such complex cases, the previous checks for {\it appropriate} representations in \gls*{cas} are insufficient. Therefore we implement numerical tests as an additional step.

@@

Considering machine accuracy and the default precision to $10$ significant digits, we can regard the first and last values as zero differences. While this evaluation is very powerful, it has a significant problem. Even when all tested values return zero, it does not prove that~\eqref{eq:branch-cut-near} was {\it appropriately} translated. However, when the values are different from zero, it does indicate that there might be an error satisfying one of the four cases~\parencite{NumericalTests:Paper}:
\begin{enumerate}
\item the numerical engine tests invalid combinations of values;
\item the translation was {\it inappropriately} defined;
\item there may be an error in the \gls*{dlmf} source; or
\item there may be an error in \Maple.
\end{enumerate}

@@

The translator was designed to be easily extendable. This allows one to implement translations for other \gls*{cas} without much effort.
However, most \LaTeX{} sources, such as arXiv, are given in generic \LaTeX{}. Semantic \LaTeX{}, which is a prerequsite for our translator, is currently prevalent in the DLMF and DRMF projects alone. Therefore, without exclusively given semantic information, the translator is not able to translate functions. Currently, we are working on mathematical information retrieval techniques which will allow for an extension of the translator to generic \LaTeX{} inputs.

>>>>

A \gls*{cas} user needs to fully understand the properties and special definitions (such as the position of branch cuts) in the \gls*{cas} to avoid mistakes during a translation~\parencite{Maple:Cuts}. A manual translation process is not only laborious, but also prone to errors. Note that this general problem has been named as automatic \gls*{p2c} conversion~\parencite{POM-Tagger}.

@@

In addition, we also introduce the notion of \textit{direct} translations. Most mathematical objects in one system have a direct counterpart in other systems. Later in the paper, we will explain that a translation from one specific mathematical object to its counterpart in the other system is not always \textit{appropriate}. Also, not every mathematical object has a counterpart in other systems. We call a translation to its counterpart \textit{direct}. For example, the translation~(\ref{eq:cos-def}) is \textit{direct}, while a translation to the definition of the cosine function
\begin{equation*}
\verb|\cos@{z}| \overset{\langMaple}{\mapsto} \verb|(exp(I*z)+exp(-I*z))/2|
\end{equation*}
is not a \textit{direct} translation even though it is {\it appropriate}.
Note that partial results of this paper have been published in~\parencite{CICM:Paper}.

\section{Related Work}\label{sec:related-work}
Since \LaTeX{} became the de facto standard for writing papers in mathematics, most \gls*{cas} provide simple functions to import and export mathematical \LaTeX{} expressions\footnote{The selected \gls*{cas} Maple, Mathematica, Matlab, and SageMath provide import and/or export functions for \LaTeX:\\
Maple, \url{http://www.maplesoft.com/support/help/Maple/view.aspx?path=latex} seen 06/2017;
Mathematica, \url{https://reference.wolfram.com/language/tutorial/GeneratingAndImportingTeX.html} seen 06/2017;
Matlab, \url{https://www.mathworks.com/help/symbolic/latex.html} seen 06/2017;
SageMath, \url{http://doc.sagemath.org/html/en/tutorial/latex.html} seen 06/2017.}. Those tools have two essential problems. They are only able to import simple mathematical expressions, where the semantics are unique. For example, the internal \LaTeX{} macro \verb|\frac| always indicates a fraction. For more complex expressions, e.g., the Jacobi polynomial in \Cref{tab:JacobiP-usecase}, the import functions fail. The second problem appears in the export tools. Mathematical expressions in \gls*{cas} are fully semantic. Otherwise the \gls*{cas} wouldn't be able to compute or evaluate the expressions. During the export process, the semantic information is lost, because generic \LaTeX{} is not able to carry sufficient semantic information. Because of these problems, an exported expression cannot be imported to the same system again in most cases (except for simple expressions such as those described above). Our tool attempts to solve these problems and provide round-trip translations between \LaTeX{} and \gls*{cas}.

@@

\vspace{-0.2cm}
The position of the branch cut of the arccotangent function differs after the {\it direct} translation~\eqref{eq:acot-alternatives}, which may lead to incorrect calculations later on. The alternative translations~\eqref{eq:acot-alternatives-1} and~\eqref{eq:acot-alternatives-2} use other functions instead of the arccotangent function. The arctangent function~\eqref{eq:acot-alternatives-1} and the natural logarithm~(\ref{eq:acot-alternatives-2}) have the same positioned branch cuts as in the \gls*{dlmf} and in \Maple. Therefore, translation~(\ref{eq:acot-alternatives-1}) solves this issue as long as the user does not evaulate the function at $z = 0$, while translation~(\ref{eq:acot-alternatives-2}) solves the issue except at $z=-\iunit$. Note that none of the translations~(\ref{eq:acot-alternatives}-\ref{eq:acot-alternatives-2}) are {\it appropriate}.

@@

The parsed tree generated by the \gls*{pom}-Tagger is not a mathematical expression tree. The \gls*{pom} project aims to disambiguate mathematical \LaTeX{} expressions and generates an expression tree. In the current state, however, many expressions still cannot be disambiguated. Therefore, the \gls*{pom}-tagger generates a raw parsed tree where each token in the \LaTeX{} expression is a node in the tree. We call this parsed tree the \gls*{pom-pt}.

@@

With these tools, we can translate a \LaTeX{} expression by translating the \gls*{pom-pt} node by node and perform group or reordering operations for some special cases. The algorithm is realized in a simple recursive structure. Whenever the algorithm finds a leaf, it can translate this single term. If the node is not a leaf, it starts to translate all children of the node recursively. This idea appears to be a practical and elegant solution, but it has a significant drawback. It cannot be used to translate functions. Since the arguments of functions are following siblings in the \gls*{pom-pt}, the algorithm needs to lookahead when a leaf is a known function, e.g., in the case of a semantic macro with arguments (see~\Cref{fig:syntax-tree-usecase}). Algorithm~\ref{alg:translation} is an improved version with lookahead functionality.

@@

Since it is more convenient to parse expressions using uniform notations, most programming 
languages (and \gls*{cas} as well) internally use prefix or postfix notation and do not mix 
the notations in one expression.
The common practice in science is to use mixed notations in expressions. Since the \gls*{pom} has rarely implemented mathematical grammatical rules, it takes the input as it is and does not build an expression tree. Therefore, it parses all four examples from~\Cref{tab:notations} to four different \gls*{pom-pt}s rather than to one unique expression tree. In general, this is not a problem for our translation process since most \gls*{cas} are familiar with most common notations. Therefore, the translator does not need to know that `$a$' and `$b$' are the operands of the binary operator `$+$' in `$a+b$.' The translator could simply translate the symbols in `$a+b$' in the same order as they appear in the expression and the \gls*{cas} would understand it. However, there are two new problems with this approach.
\begin{enumerate}\item \label{prob:1} The translated expression is only syntactically correct if the input expression was syntactically correct.
\item \label{prob:2} We cannot translate expressions to \gls*{cas} which use non-standard notations.
\end{enumerate}

@@

\subsection{Subtranslators}\label{sec:subtranslators}
The \verb|SequenceTranslator| translates the \textit{sequence} and \textit{balanced expressions} in the \gls*{pom-pt}. If a node $n$ is a leaf and the represented symbol is an open bracket (parentheses, square brackets and so on), the following nodes are also taken as a \textit{sequence}. Combined with the recursive translation approach, the \verb|SequenceTranslator| also checks balancing of parentheses in expressions. An expression such as `$(a]$' produces a mismatched parentheses error. On the other hand, this is a problem for real interval expressions such as `$[a,b)$'. In the current version, the program cannot distinguish between mismatched parentheses and half-opened, half-closed intervals. Whether an expression is an interval or another expression is difficult to decide and can depend on the context. Also, the parentheses checker could simply be deactivated to allow mismatched parentheses in an expression. Another option is to use interval macros. e.g., \verb|\intcc@{a}{b} = [a,b]|.

@@

There are two essential problems with this approach. Testing whether expressions are {\it appropriate} representations of each other is a challenging task for \gls*{cas} and they often have difficulties testing simple equations symbolically. For example, 
consider
\parencite[(4.35.34)]{NIST:DLMF}
\begin{equation*}
\sinh \left( x+\iunit y \right) = \sinh x \cos y + \iunit \cosh x \sin y,
\end{equation*}
as a difference of the left- and right-hand sides cannot be simplified to zero by default. Furthermore, this approach only checks forward translations because there is no way to automatically check whether two \LaTeX{} expressions are {\it appropriate} or {\it inappropriate} representations of each other (again this could become feasible with our translator). We use \Maple's \texttt{simplify} function to check if the difference of the left-hand side and the right-hand side of the equation is equal to zero. In addition, we use \texttt{simplify} and check if the division of the right-hand side by the left-hand side returns a numerical value or not. The simplification function is the most powerful function to check whether two expressions are {\it appropriate} representations in \Maple. However, there are several cases where simplification fails. Because of implementation details, there are some techniques that help \Maple{} to find possible simplifications. For example, we can force \Maple{} to convert the formula
\begin{equation}
\sinh{x} + \sin{x}
\end{equation}
to an {\it appropriate} representation using their exponential representations, namely
\begin{equation}
\frac{1}{2}\expe^x - \frac{1}{2}\expe^{-x} - \frac{1}{2} \iunit \left( \expe^{\iunit x}-\expe^{-\iunit x} \right).
\end{equation}
With such pre-conversions, we are able to improve the simplification process in \Maple. However, the limitations of the \texttt{simplify} function are still the weakest part of this verification approach. Consider the complex example~\parencite[(12.7.10)]{NIST:DLMF}
\begin{equation}\label{eq:branch-cut-near}
\displaystyle U(0,z) = \sqrt{\frac{z}{2\cpi}} K_{\frac{1}{4}}\left(\frac{1}{4}z^2\right),
\end{equation}
where $U(0,z)$ is the parabolic cylinder function and $K_\nu(z)$ is the modified Bessel function of the second kind. Both functions are well-defined in both systems and we can define a \textit{direct} translation for~\eqref{eq:branch-cut-near}. 
The modified Bessel function of the second kind has its branch cut in \Maple{} and in the \gls*{dlmf} at $z < 0$. However, the argument of $K$ contains a $z^2$. If $|\ph{z}| \in \left(\frac{\cpi}{2}, \cpi\right)$ the value of the right-hand side of~\eqref{eq:branch-cut-near} would be no longer on the principal branch. \Maple{} will still compute the principal values independently of the value of $z$ and the translation
\begin{equation}
\verb|\BesselK{\frac{1}{4}}@{\frac{1}{4}z^2}| \overset{\langMaple}{\mapsto} \verb|BesselK(1/4,(1/4)*z^2)|
\end{equation}
is {\it inappropriate} if $|\ph{z}| \in \left(\frac{\cpi}{2}, \cpi\right)$. One should instead use the analytic continuation for the right-hand side of (\ref{eq:branch-cut-near}). 
To evaluate such complex cases, the previous checks for {\it appropriate} representations in \gls*{cas} are insufficient. Therefore we implement numerical tests as an additional step.

@@

Considering machine accuracy and the default precision to $10$ significant digits, we can regard the first and last values as zero differences. While this evaluation is very powerful, it has a significant problem. Even when all tested values return zero, it does not prove that~\eqref{eq:branch-cut-near} was {\it appropriately} translated. When the values are different from zero, it does indicate that there might be an error satisfying one of the four cases~\parencite{NumericalTests:Paper}:
\begin{enumerate}
\item the numerical engine tests invalid combinations of values;
\item the translation was {\it inappropriately} defined;
\item there may be an error in the \gls*{dlmf} source; or
\item there may be an error in \Maple.
\end{enumerate}

@@

The translator was designed to be easily extendable. This allows one to implement translations for other \gls*{cas} without much effort.
However, most \LaTeX{} sources, such as in arXiv, are given in generic \LaTeX{}. Semantic \LaTeX{}, which is a prerequsite for our translator, is currently prevalent in the DLMF and DRMF projects alone. Therefore, without exclusively given semantic information, the translator is not able to translate functions. Currently, we are working on mathematical information retrieval techniques which will allow for an extension of the translator to generic \LaTeX{} inputs.
